\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-06 chapters/bttda.tex
cover page

\input{chapters/bttda/tensorviz.tex}

@@

@@

If the decompositions are not full rank, the Tucker, PARAFAC and block-term
structures are not unique and can be obtained by optimizing different criteria.
Given the low signal-to-noise ratio and specific, task-related output expected
in a BCI application, supervised feature extraction and machine learning techniques are
favored~\cite{Lotte2018} over the unsupervised decomposition methods presented
above.
A decomposition that is helpful for classification purposes should ideally optimize
the discriminability between classes in the resulting core tensors, which can
be considered as extracted features.
In this philosophy, the Tucker decomposition can also be obtained
using Higher Order Discriminant Analysis
(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fischer sense, analogous to linear
discriminant analysis.
Extracted features can subsequently be further classified, most commonly
using LDA or a support vector machine (SVM) to predict class labels.

@@

@@

\subsection{A forward model for HODA}
As a prerequisite to the proposed BTTDA model, we must find a
way to reconstruct the original data tensor $\ten{X}$ as accurately as possible
from $\ten{G}$ after dimensionality reduction.
This is usually referred to as finding a corresponding \emph{forward model}.
While a \emph{backward model} extracts latent sources or properties from the observed
data based on some task related criterion or on prior domain knowledge,
a forward model is a generative model that expresses the observed data in
function of these latent properties or sources that are given.
Forward models are useful for a.o.\ interpretability and data compression.
\todo{explicitly label error terms in equation, use $X\prime$ and error with
iteration sup}

The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model is given by
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$ and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA.
    By calculating activation	patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}
\todo{not referred to in text}
\todo{I would prefer a figure in which the Ai and G are on the left and
$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
use a standard equal signal. Finally, in the caption, you could clearly define
the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
\todo{explicitly state why choosing this versus full model or HOPLS}

A forward model should make sure that reconstruction error is minimized and
variation captured in the latent tensor is maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
\todo{not clear ("error is minimized" in the same sentence)}
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
  -\hat{\ten{X}}(n)\right)^2
  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}
\todo{Reorder figures and add error term}

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
Equation~\ref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
\todo{eps1 = X - G1 x {A1} still contains discriminative info, so
it is further projected onto another core tensor yielding
eps2 = eps1 - G2 x
{A2}
etc. and finally you give equation (15)
}
We thus extend the HODA feature extraction scheme with backward and
forward models defined in respectively Equations~\ref{eq:HODA-backward}
and~\ref{eq:HODA-forward} to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
Figure~\ref{fig:BTTDA} further illustrates the BTTDA model.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
  \caption[A forward model for \ac{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
\todo{Explicitly mention that if restricted to rank one, it is parafac, and to
one block, it is hoda}

Since BTTDA is specified as a forward model, a backward modelling
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
  \caption{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \todo{ref equation} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

Furthermore, HODA, and by extension BTTDA can extract a substantial amount
of redundant features, which can be dropped after projection and before proceeding to the classification
step~\cite{Phan2010}.
Specifcally in BTTDA redundant features can stack up over the number of
blocks, hampering performance.
Relevant features can be retained by calculating the
univariate Fisher score for all features.
The Fisher score $\phi{i}$ for feature $i$ is obtained as
\begin{equation}
	\phi(i) = \frac
	{\sum_c^C N_c \left(\bar{g}_i(c)-\bar{\bar{g}}_i\right)^2}
	{\sum_n^N \left(g_i(n)-\bar{g}_i(c_n)\right)^2}
\end{equation}
The features can then either be sorted by $\phi(i)$ and a given amount of
features retained for classification, or the decision to retain a feature can
be made by a threshold on the statistical $p$-values corresponding to the
$\phi(i)$.
We opted for the latter strategy since it does not require us to redetermine the
optimal number of features to retain at each block.

@@

\subsection{Block contribution}
\todo{Discuss number of features}
To analyze the contribution of extra feature blocks extracted by {BTTDA} over
the first one found by HODA we pick n ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation and test data as a function
of the number of blocks increases, Figure~\ref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\ten{X}_\text{rec}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as
\begin{equation}
	\nmse\left(\ten{X}, \ten{X}_\text{rec}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
	\input{figures/bttda/blocks.pgf}
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

\section{Discussion}
\todo{subtensors in larger tensor drawing, focus more on sparsity in the
discussion}
\todo{focus on tradeof: more flexibility in model, but this implies more
hypermarameters so it relies more on a model selection procedure.}
The results in Table~\ref{tab:erp-score} and Table~\ref{tab:mi-score} show that
in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
exceeding all comparison methods for the two evaluated datasets, but
for MI datasets results fall short of those of comparison methods.
Performance on the MI classification task could however be greatly influenced
by the exact tensorization method used, i.e. the time-frequency transform in
this case, which is not the main focus of this work.

@@

We assume the main benefit of BTTDA is that it can more easily discover relevant
features, while
being more parsimonious due to its block-term structure compared to HODA's full
Tucker structure.
Alternatively, the enhanced performance could also stem from the modeled data
covariance.
Since HODA estimates one whithin-class scatter matrix
$S_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
model of the data scatter is determined by these per mode scatter matrices as a
Kronecker product $S_{-1,\text{w}}\otimes S_{-2,\text{w}}\otimes\cdots\otimes S_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
drawn from a multilinear normal distribution~\cite{Ohlson2013}.
However, it is known that the EEG covariance cannot fully be expressed as a
single Kronecker product, but rather is more accurately modeled by a sum of
multiple Kronecker products~\cite{Bijma2005, Sosulski2022}.
Since BTTDA iteratively fits HODA models to the residual error, it will be able
to express the full covariance structure given sufficient blocks.

@@

Despite favorable results in BCI decoding, the applications of the proposed
BTTDA model are limited mainly by the model selection approach used
to determine the individual block ranks.
While our proposed greedy model selection algorithm is a step in the right
direction, the high computational cost of setting hyperparameters through cross
validation can still hinder the portability of decoders relying on
BTTDA.
Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal between- or within-block feature
correlation, decreasing discriminability etc.
Future efforts should focus on automatic parameter setting, e.g. using
information criteria such as the ones used in BTTR~\cite{Faes2022} or other
statistical measures based on the models application.
\todo{comment as limitation that we test r1=r2=r3, this is against the whole
idea of increased flexibility, find automated rank selection procedure}

Another limitation is that BTTDA might yield a disproportionate
improvement for datasets with a low number of features relative to sample size,
while being less effective for datasets with more features
This is reflected
in our ERP results (low dimensionality vs. high number of trials) compared to the
MI results (higher dimensionality due to third order tensorization vs. lower
number of trials).
We expect a dimensionality limit beyond which the forward modeling step cannot
accurately regress from the low dimensional latent tensors to the high
dimensional original tensors, introducing
error in the input data for the next block, which can stack up over blocks.
Since the forward multilinear least squares problem is underdetermined, it is prone to
numerical instability, which calls for regularization of the forward modeling
procedure, but this would introduce another hyperparameter.
It should also thoroughly be investigated what the impact is of going beyond
second- and third-order case to higher order tensors, since this could have a large impact
on the model.
Other tensorization methods of the EEG data, like time-lagged Hankel
tensors~\cite{Papy2005}, or tensors across subjects or sliding windows etc., could also be
of interest if they are appropriately chosen based on prior knowledge of the dataset.
\todo{mention in discussion of model selection procedure that sparser blocks
would be better}

\section{Conclusion}
\todo{main restriction for hoda is model selection. this is also a model
selection problem, but it is more flexible}
We have introduced Block-Term Tensor Discriminant Anlysis (BTTDA) a novel,
tensor-based, supervised dimensionality reduction technique optimized for class
discriminability which adheres to the block-term tensor structure.
BTTDA is a generalization of Higher Order Discriminant Analysis and can also be
applied as a special sum-of-rank-one tensors PARAFACDA model.
The model is obtained by iteratively fitting HODA in a deflation scheme,
leveraging a novel forward modeling step.
Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
feature extraction can significantly outperform decoders based on HODA and
reach state-of-the art decoding performance on event-related potential
problems (second order tensors) and scores on par with or higher than HODA  in motor
imagery problems (third order tensors).
Moving from the rigid Tucker tensor structure of HODA to the more flexible
block-term structure shifts the problem from finding optimally constrained multilinear
projections to model and feature selection.
This allows performance to be traded
off for model complexity and number of features, to find a setting that is more
effective for decoding.
Because of the flexibility and minimal assumptions BTTDA
can equally be applied to other neuroimaging modalities (MEG, ECoG, fNIRS,
fMRI, EMG, ...) or tensor classification problems in other fields.

\section*{Acknowledgements}
\todo{move to the end}
We thank the Flemish Supercomputer Center (VSC) and the High-Performance
Computing (HPC) center of the KU Leuven for allowing us to execute our
computational experiments on their systems.
We also wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing this
work.

>>>>

\chapter{Block-term tensor discriminant analysis}
\label{sec:bttda}
\input{chapters/bttda/tensorviz.tex}
\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}}.

@@

\subsection{Tensor \& tensor methods}

@@

\subsection{Supervised tensor decomposition}

If the decompositions are not full rank, the Tucker, PARAFAC and block-term
structures are not unique and can be obtained by optimizing different criteria.
Given the low signal-to-noise ratio and specific, task-related output expected
in a BCI application, supervised feature extraction and machine learning techniques are
favored~\cite{Lotte2018} over the unsupervised decomposition methods presented
above.
A decomposition that is helpful for classification purposes should ideally optimize
the discriminability between classes in the resulting core tensors, which can
be considered as extracted features.
In this philosophy, the Tucker decomposition can also be obtained
using Higher Order Discriminant Analysis
(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fisher sense, analogous to linear
discriminant analysis.
Extracted features can subsequently be further classified, most commonly
using LDA or a support vector machine (SVM) to predict class labels.

@@

\subsection{A block-term structured model for classification}

@@

\subsection{A forward model for HODA}
As a prerequisite to the proposed BTTDA model, we must find a
way to reconstruct the original data tensor $\ten{X}$ as accurately as possible
from $\ten{G}$ after dimensionality reduction.
This is usually referred to as finding a corresponding \emph{forward model}.
While a \emph{backward model} extracts latent sources or properties from the observed
data based on some task related criterion or on prior domain knowledge,
a forward model is a generative model that expresses the observed data in
function of these latent properties or sources that are given.
Forward models are useful for a.o.\ interpretability and data compression.

The HODA projection in Equation~\ref{eq:HODA-backward} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model, visualized in Figure~\ref{fig:HODA-forward} is given by
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$ and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA.
    By calculating activation	patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}

A forward model should make sure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
  -\hat{\ten{X}}(n)\right)^2
  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

The forward model could alternatively be specified in the classical manner by linear
regression or with the solution of \textcite{Haufe2014}, but this would require
vectorizing the tensor representation.
While the reconstruction error might be brought down, we lose the computational
efficiency and regularizing constraints of the tensor form.
More efficiently, the forward model could be obtained through other tensor
regression methods, like Higher Order Partial
Least Squares.
This method still requires fitting more parameters than our
proposed model, on top of additional hyperparameters defining the
dimensionality of the common subspace.
An attractive property of our multilinear forward model for HODA is that it
estimates exactly as many parameters as the backward model, with the intuition that
it should not disproportionally contribute to overfitting.

\subsection{Block-Term Tensor Discriminant Analysis}
After defining the forward model, we can construct our proposed block-term
tensor model.
Assuming the latent tensors $\ten{G}$
obtained by the backward projection of HODA do not achieve perfect
class separation, the error term $\ten{\mathbfcal{E}}$ in
Equation~\ref{eq:HODA-forward} should still contain some discriminative
information, which can be exploited to improve classifier
performance.
Useful features can still be extracted from $\ten{\mathbfcal{eps}} = \ten{X} -
\hat{\ten{X}}$, so it is further projected onto another core tensor
$\ten{G}^{(2)}$ (assuming $\ten{G}$ as $\ten{G}^{(1)}).$

We thus extend the HODA feature extraction scheme to Block-Term Tensor Discriminant Analysis
(BTTDA).
BTTDA finds multiple discriminative blocks, such that its forward
model adheres to the block-term tensor structure:
\begin{align}
	\ten{X} & = \sum_b^B\ten{G}^{(b)}\mmpr{\mat{A}^{(b)}} + \ten{\mathbfcal{E}}
	\label{eq:BTTDA-forward}
\end{align}
for $B$ extracted latent tensors $\ten{G}^{(b)}$ and residual error term
$\ten{\mathbfcal{E}}$.
Figure~\ref{fig:BTTDA} further illustrates the BTTDA model.
\begin{figure*}[t]
	\centering
	\input{figures/bttda/BTTDA.tikz.tex}
  \caption[A forward model for \ac{bttda}.]{A forward model for Block-Term Tensor Discriminant Analysis
		(BTTDA). BTTDA can extract more features
		than HODA by iteratively finding a latent tensor $\ten{G}^{(b)}$ in a
		deflation scheme.
		The HODA backward projection is first applied. Next, the
		input data is reconstructed via the HODA forward model and the
		difference between the two is found.
		Finally, this process is repeated with this difference as input data, until a
		desired number of blocks $B$ has been found.}
	\label{fig:BTTDA}
\end{figure*}
The block-term structure of this model implies that is a generalization of both
the Tucker-structured HODA and PARAFAC-structured discriminant feature
extraction.
If $B$ in Equation~\ref{eq:BTTDA-forward} is set to one, BTTDA is equivalent to
HODA; if at each term $b$ the rank of the core tensor
$(r_1^{(b)}=r_2^{(b)}=\ldots=r_k^{(b)}=1)$, a PARAFAC structure is assumed.

Since BTTDA is specified above as a forward model, a backward procedure
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
  \caption{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in Equition~\ref{eq:HODA-backward} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

Furthermore, HODA, and by extension BTTDA can extract a substantial amount
of redundant features, which can be dropped after projection and before proceeding to the classification
step~\cite{Phan2010}.
Specifcally in BTTDA redundant features can stack up over the number of
blocks, hampering performance.
Relevant features can be retained by calculating the
univariate Fisher score for all features.
The Fisher score $\phi(i)$ for feature $i$ is obtained as
\begin{equation}
	\phi(i) = \frac
	{\sum_c^C N_c \left(\bar{g}_i(c)-\bar{\bar{g}}_i\right)^2}
	{\sum_n^N \left(g_i(n)-\bar{g}_i(c_n)\right)^2}
\end{equation}
The features can then either be sorted by $\phi(i)$ and a given amount of
features retained for classification, or the decision to retain a feature can
be made by a threshold on the statistical $p$-values corresponding to the
$\phi(i)$.
We opted for the latter strategy since it does not require us to redetermine the
optimal number of features to retain at each block.

@@

\subsection{Block contribution}
\todo{Discuss number of features}
To analyze the contribution of extra feature blocks extracted by {BTTDA} over
the first one found by HODA we pick n ERP ($K=2$) dataset
(BNCI2014-008) and an MI ($K=3$) dataset (BNCI2014-001).
We report within-session ROC-AUC scores for training, validation and test data as a function
of the number of blocks increases, Figure~\ref{fig:blocks}.
Training and validation folds were taken from the model selection procedure.
Additionally, the Normalized Mean Squared Error (NMSE) is reported for the
reconstructed from the truncated BTTDA decomposition
$\textstyle{\hat{\ten{X}}^{(B)}=\sum_b^B\ten{G}^{(b)}\mmpr{\mat{U}^{(b)}}}$.
NMSE is calculated as
\begin{equation}
  \nmse\left(\ten{X}, \hat{\ten{X}}^{(B)}\right) =
	\frac{\sum_n^N\left\lVert\ten{X}(n)-\ten{X}_\text{rec}^{(B)}(n)\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{X}(n)\right\rVert_F^2}
\end{equation}
\begin{figure*}[t]
	\input{figures/bttda/blocks.pgf}
  \caption[Analysis of NMSE and classification score per block.]{%
    Normalized Mean Square Error (NMSE) (left), and difference in
    classification score for the training and validation for the ERP datasets
    (top row) and the MI dataset (bottom row)	of the greedy model selection
    procedure (right), as a function of the number of BTTDA blocks $b$.
		While NMSE monotonically decreases for the evaluated datasets, better class
		separation will be achieved, but eventually overfitting occurs and validation
		and test scores will drop, or plateau due to feature selection.
	}
	\label{fig:blocks}
\end{figure*}
\todo{format figure}

\section{Discussion}
\subsection{Contribution}
The results in Table~\ref{tab:erp-score} and Table~\ref{tab:mi-score} show that
in the ERP datasets, BTTDA+LDA and PARAFACDA+LDA reach state-of-the-art decoding performance by
exceeding all comparison methods for the two evaluated datasets, but
for MI datasets results fall short of those of comparison methods.
Performance on the MI classification task could however be greatly influenced
by the exact tensorization method used, i.e. the time-frequency transform in
this case, which is not the main focus of this work.

@@

We assume the main benefit of BTTDA is that it can more easily discover relevant
features, while
being more parsimonious due to its block-term structure compared to HODA's full
Tucker structure, as illustrated by Figure~\ref{fig:bttda/sparse}.
\begin{figure}
  \input{figures/bttda/sparse.tikz.tex}
  \caption[Sparsity of BTTDA and PARAFACDA]{BTTDA and PARAFACDA can find a
  sparser expression of the discriminant information captured by the dense,
  Tucker-structured HODA algorithm.}
  \label{fig:bttda/sparse}
\end{figure}
The same discriminative information captured by a relatively large
Tucker-structured core tensor, could be expressed more sparsely with a small
number of block-terms, while avoiding redundant features.
The PARAFAC structure employed in PARAFACDA is even more sparse, which could be
a benefit or an advantage depending on the amount of regularization required,
or on the true underlying structure of the data.

Alternatively, the enhanced performance could also stem from the modeled data
covariance.
Since HODA estimates one whithin-class scatter matrix
$S_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
model of the data scatter is determined by these per mode scatter matrices as a
Kronecker product $S_{-1,\text{w}}\otimes S_{-2,\text{w}}\otimes\cdots\otimes S_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
drawn from a multilinear normal distribution~\cite{Ohlson2013}.
However, it is known that the EEG covariance cannot fully be expressed as a
single Kronecker product, but rather is more accurately modeled by a sum of
multiple Kronecker products~\cite{Bijma2005, Sosulski2022}.
Since BTTDA iteratively fits HODA models to the residual error, it will be able
to express the full covariance structure given sufficient blocks.

@@

\subsection{Model selection and dimensionality}

BTTDA trades in the rigid HODA model for increased model complexity with more
hyperparameters to tune, which could open up a setting where performance can be
improved.
This turns tensor discriminant analysis into a model selection procedure
instead of a projection optimization procedure.
Despite favorable results in BCI decoding, the applications of the proposed
BTTDA model are limited mainly by this model selection approach used
to determine the individual block ranks.
While our proposed greedy model selection algorithm is a step in the right
direction, the high computational cost of setting hyperparameters through cross
validation can still hinder the portability of decoders relying on
BTTDA.

Due to its heuristic nature, the greedy algorithm does not always result in the
set of ranks with the highest achievable performance.
In combination with the fact that the feature selection cutoff parameter was
fixed somewhat arbitrarily, it is clear that thorough
hyperparameter optimization could improve performance.
Furthermore, it is clear that our proposed model selection procedure does not
necessarily result in an optimal set of blocks that group coherent projections
within the same block, according to some some desirable metric.
Examples of this are sparsity, pattern interpretability, minimal or maximal between- or within-block feature
correlation, decreasing discriminability etc.
Finally, the presented results are limitation that by our artificial restriction
that $(r_1=r_2=\ldots=r_k)$ to reduce computational demand of the presented
experiments.
In a sense, this goes against the proposition of increased model flexibility.
It might well be that BTTDA in some cases offers little added value over the
Tucker-structured HODA, when the both are given free choice of rank.
Future efforts should focus on automatic parameter setting, e.g. using sparsity
or information criteria such as the ones used in BTTR~\cite{Faes2022} or other
statistical measures based on the models application.

Another limitation is that BTTDA might yield a disproportionate
improvement for datasets with a low number of features relative to sample size,
while being less effective for datasets with more features
This is reflected
in our ERP results (low dimensionality vs. high number of trials) compared to the
MI results (higher dimensionality due to third order tensorization vs. lower
number of trials).
We expect a dimensionality limit beyond which the forward modeling step cannot
accurately regress from the low dimensional latent tensors to the high
dimensional original tensors, introducing
error in the input data for the next block, which can stack up over blocks.
Since the forward multilinear least squares problem is underdetermined, it is prone to
numerical instability, which calls for regularization of the forward modeling
procedure, but this would introduce another hyperparameter.
It should also thoroughly be investigated what the impact is of going beyond
second- and third-order case to higher order tensors, since this could have a large impact
on the model.
Other tensorization methods of the EEG data, like time-lagged Hankel
tensors~\cite{Papy2005}, or tensors across subjects or sliding windows etc., could also be
of interest if they are appropriately chosen based on prior knowledge of the dataset.

\section{Conclusion}

We have introduced Block-Term Tensor Discriminant Anlysis (BTTDA) a novel,
tensor-based, supervised dimensionality reduction technique optimized for class
discriminability which adheres to the block-term tensor structure.
BTTDA is a generalization of Higher Order Discriminant Analysis and can also be
applied as a special sum-of-rank-one tensors PARAFACDA model.
The model is obtained by iteratively fitting HODA in a deflation scheme,
leveraging a novel forward modeling step.

Via an accompanying heuristic model selection procedure, BCI decoders using BTTDA
feature extraction can significantly outperform decoders based on HODA and
reach state-of-the art decoding performance on event-related potential
problems (second order tensors) and scores on par with or higher than HODA  in motor
imagery problems (third order tensors).
Moving from the rigid Tucker tensor structure of HODA to the more flexible
block-term structure shifts the problem from finding optimally constrained multilinear
projections to model and feature selection.

Introducing a flexible block-term tensor model as underlying structure
reformulates tensor discriminant analysis as a model selection
problem.
This allows performance to be traded off for model complexity and number of
features, to find a setting that is more effective for decoding.
Because of its general implementation and minimal assumptions on data structure,
BTTDA can equally be applied to classification for other neuroimaging modalities
(MEG, ECoG, fNIRS, fMRI, EMG, ...) or to tensor classification problems in other
fields.
